\chapter{The cover tells the story}
\vspace{-2cm}
\begin{center} 
by Nikyta Guleria
\end{center}
\vspace{2cm}


I designed this cover to capture the essence of the book; a book embedded with ideas that were taking form long before Graham and I began working together six years ago. Excited to take on the challenge, this book was more than an omnibus of stories that gave me hope and spelled optimism. It was a medley of practical tools, researched methodologies, and voices of reason, captured in a way that left me visualizing a vibrant cohesive world. It also gave me the ability to recognize the gray inadequacies of today’s world, ones that bar us from meeting large-scale change and transformation. It is this contrast between today’s ‘old world’ and tomorrow’s ‘new world’ that sits center stage in the cover artwork.  


In order to rebuild without risking catastrophic collapse, we must simultaneously build the new and dismantle the old world, without losing the life support of the old, nor succumbing to the toxicity of the old world (denoted in gray undertones) dominating the lives that most of us still live. A world that fails almost everyone on the planet today; and prevents the next generations from thriving. A world that is slowly but surely showing its cracks. 


This world’s systems and structures systemically exclude, or discriminate against, people who would love to contribute to a thriving society; sometimes driving them to the fringes of society. Capturing the physically and psychologically dangerous elements rooted in our workplace practices and cultures, at best, I was able to share glimpses of what this waste of human lives, potential, and resources looks like, seen in the form of burn-out, unrest, depression, homelessness, and even loss of life.


Step by step we need to dismantle our linear use\hyp{}and\hyp{}dump mentality and practices (depicted through deforestation, garbage pileups, and demolitions), dismantle everything putting toxic waste into nature and our communities (oil well, factory chimneys, and fumes), and replacing them with circular, fully regenerative practices.


We live in this very fading, graying old world, that often blocks us from seeing the vibrant, regenerative world we ought all to thrive in (depicted in bright hues of blue, green and purple). This new world doesn’t exist yet and we cannot know what it will
look like, nor can we know all the elements we will need to build it. But we do
know the regenerative, life\hyp{}giving essence it must have (depicted as individual and overlapping elements moving in harmony). In order to see this new world unfold, it will need us to look through new lenses, just as Picasso and the other founders of cubism did. Look for the eyes and faces in the new world, representing the capacity we most need now: to see what others cannot yet see, to imagine possibilities others cannot yet imagine, and to build what others cannot yet build.


The book beautifully articulates a connection between the ‘economy’, ‘leadership’ and ‘you’, as a constant, sustainable, even regenerative motion, like the cyclic nature of waves and tides of our oceans and atmosphere or the ‘waves’ of seasons we see moving at a glacial speed, shaping our landscape. A constant re\hyp{}generative motion that is conducive to shaping everything, connecting everything, and vital for moving energy and resources to where life needs it.


As I imagined this world together with the authors, the image in our minds was dominated by colourful waves representing building blocks that are still to be defined; hope, flow, celebrations, the visible and the hidden, opportunities and threats, emotions, adventures, successes, failures, and everything in between and neither; and much more. They represent the opportunities we have right now to ride the tide, as Brutus points at in Shakespeare’s Julius Caesar:
\begin{quotation}
There is a tide in the affairs of men \newline
Which, taken at the flood, leads on to fortune. 
\end{quotation}


In the midst of complex and chaotic legacy ways of living and operating, Graham and Jack, through their words, have managed to paint a clear and pure picture of what tomorrow could look like, by connecting the dots and highlighting the red threads that string together our conscious and unconscious worlds, clearing the path for individuals and organizations to take action and create a future befitting for humanity in the 21st century. 




\begin{comment}




I designed the cover to tell the story of the book, a story Graham has been telling since long before I began working with him, six years ago. A story that still gives me hope, a story of how we can realistically build a new world today, a practical world that works for everyone and all life on spaceship Earth. 


A new world filled with constant, sustainable, even regenerative motion, like the waves and tides of our oceans and atmosphere that we can see every day; the seasons, that we see over months; and the waves shaping our landscape, that we can only see in geological records. The constant regenerative motion shaping everything, connecting everything, moving energy and resources to where life needs it. 


So the image of the new world is dominated by colourful waves representing building blocks that are still to be defined; hope, flow, celebrations, the visible and the hidden, opportunities and threats, emotions, adventures, successes, failures, and everything in between and neither; and much more. They represent the opportunity we have \emph{right now} to ride the tide Brutus points at in Shakespeare’s Julius Caesar: \begin{quote}There is a tide in the affairs of men \newline Which, taken at the flood, leads on to fortune.\end{quote}


This new world, though, doesn’t exist yet. We cannot yet know what it will look like, nor can we know all the elements we will need to build it. But we do know the regenerative, life-giving essence it must have, and that to see it we need to look through new lenses, just as Picasso and the other founders of cubism did. Look for the eyes and faces in the new world, representing the capacity we most need now: to see what others cannot yet see, to imagine possibilities others cannot yet imagine, and to build what others cannot yet build. 


As we rebuild, we also need to dismantle, step by step, the grey old world dominating the lives of most of us still. A world that fails almost everyone on the planet today; and even worse, sets up the next generations for misery, by wasting,  in a few decades, the critical resources our descendants ought to still have access to. A world that is holding us, and the next generations, back from thriving.


This world’s systems and structures systemically exclude, or discriminate against, people who would love to contribute to a thriving society; sometimes driving them into homelessness or worse. They are psychologically dangerous, and even physically dangerous; workplace practices and cultures that, at best, waste lives and our potential to create a better world, and at worst end in burn-out, depression, and even death. 


Step by step we need to dismantle our linear use\hyp{}and\hyp{}dump mentality and practices (depicted in piles of garbage and demolitions), dismantle everything putting toxic waste into nature and our communities (oil well, factory chimneys and fumes) replacing them with circular, fully regenerative practices. 


We live in a fading,  graying old world, on a wall blocking us from seeing, let alone breaking through to, the vibrant, regenerative world we ought all to thrive in. 


We can rebuild everything we have today, so that we are all freed up to be our best, unique selves, and all thrive together. 




\end{comment}